\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-07-02 lang-design.tex
Going through before metatheory for small-fixes.

We want our metatheory mechanization to allow adding new function/proving new lemmas, and adding new constructors for inductive type at the same time  (Expression Problem~\cite{wadler-ep}). Coq support the former but not the later---because when an existent inductive type is extended with new constructor, we have to modify the pattern matching in the existent functions. This motivates us to add inheritance/Family Polymorphism into Coq.

@@

However, to add inheritance into a dependent type system like Coq is not simple and there are various challenges and tension we need to resolve.

\textlabel{Challenge (1)}{chg:extensible-inductive-type}~\textbf{Extensible Inductive Type vs. Exhaustiveness of Inductive Reasoning}. Unlike most general purposed programming languages, Coq, as a dependent type theory, is a total language and thus each pattern matching cases must fully ``cover'' an inductive type. This process is \textit{dynamic} for inheritance---when inhertiance extend the inductive type in the children family but the user forget to extend the corresponding inductive reasoning, our compiler need to signal an error correspondingly, or better---guide the user to fill the incomplete inductive reasoning. For example, in \cref{fig:STLC-example} when "term" is extended with "tt" in the children family "STLC_bool", our compiler need to make sure "subst" and "subst_lemma" are also extended correctly otherwise signaling an error.
Additionally, our compiler must support direct inheritance (and overriding) of the old case handlers, to waive the boilerplate code.

\textlabel{Challenge (2)}{chg:definition-relevant-reasoning}~\textbf{Definition-awared Reasoning vs. Overridability}. In family polymorphism for general purposed programming language, overriding is allowed to happen anywhere. But the problem becomes much more subtle in the case of dependent type---the later fields can have a type dependent on the former fields! Consider a small example where we have consecutive fields \mintinline{Coq}{{.. a : nat := 1; a-def : Eq a 1 := eq_refl; ..}}, should we support override "a"? If we do support it, how can our compiler detect that "a-def" is not inheritable and signal an error when such thing happens?

@@

These two examples both show the incompatibility between exposing the concrete definition of a field for reasoning (this second example even involves the definition of pattern matching) and overridability. We need to be careful about this tension when designing the language.

\textlabel{Challenge (3)}{chg:consistency}~\textbf{Consistency vs. Self Parametrization}. The metatheoretical formulation in~\citet{zm2017} for family polymorphism is very natural as they consider extension of the family as the concatenation of a new linkage. However, this formulation is not directly adaptable for dependent type theory because it can easily encode non-termination sabotaging consistency by the presence of "self". Informally, in the orignal formulation, we can override the "f" in family $\{"f" := id; "g" := "call self.f"\}$ with "call self.g", resulting the mutual calling of "f" and "g" thus a deadloop. This conflict between self parametrization and consistency will be another challenge---we need change the original formulation of family and inheritance for dependent type theory.

\textlabel{Challenge (4)}{chg:software-engineering}~\textbf{Engineering and Ecosystem}.  Not only for the metatheory, we also want our implementation of family to have good compatibility with the rest of Coq. For example, we cannot compromise \textbf{tactic programming} as it is already a widely accepted part of Coq programming. Additionally, the interactive development style requires \textbf{incremental type checking}---our compiler has to be able to type check each Coq Vernacular command. What's more, we want our family ``term'' to incorporate Coq \textbf{without breaking its other functionality}. For example, the user may want to project the family member anywhere inside an arbitrary Gallina term, that is later being computed and normalized. However, families themselves can be second class citizens just like \citet{zm2017}.

\YZ{I was thinking if the paper should start by talking about the functional encoding of inheritance (namely exposing a self parameter to allow late binding, and later taking the fixpoint of the extensible module when it needs to be accessed outside the family).  I think most readers will benefit from a recap of this encoding.\\
Coq is total but that doesn’t prevent Coq from having a Fixpoint keyword (plus you proved your language is consistent).
}

>>>>

We want the metatheory mechanization to allow adding new function/proving new lemmas, and adding new constructors for inductive type at the same time  (Expression Problem~\cite{wadler-ep}). Coq support the former but not the later---when an existent inductive type is extended with new constructor, we have to modify the pattern matching in the existent functions. This motivates us to add inheritance/Family Polymorphism into Coq as a solution to the Expression Problem.

@@

Refering to the syntax of \citet{zm2017}, we will come up the above example. However, to add inheritance into a dependent type system like Coq is not simple and there are various challenges and tension we need to resolve.

\textlabel{Challenge (1)}{chg:extensible-inductive-type}~\textbf{Extensible Inductive Type vs. Exhaustiveness of Inductive Reasoning}. Unlike most general purposed programming languages, Coq, as a dependent type theory, is a total language and thus each pattern matching cases must fully ``cover'' an inductive type. This process is ``\textit{interactive}'' for inheritance---when inhertiance extend the inductive type in the children family but the user forgets to extend the corresponding inductive reasoning, our compiler need to signal an error correspondingly, or better---guide the user to fill the incomplete inductive reasoning. For example, in \cref{fig:STLC-example} when "term" is extended with "tt" in the children family "STLC_bool", our compiler need to make sure "subst" and "subst_lemma" are also extended correctly otherwise signaling an error.
Additionally, our compiler must support direct inheritance (and overriding) of the old case handlers for "var", "lam" and etc, to avoid the boilerplate code.

\textlabel{Challenge (2)}{chg:definition-relevant-reasoning}~\textbf{Definition-awared Reasoning vs. Overridability}. In family polymorphism for general purposed programming language, overriding is allowed to happen anywhere. But the problem becomes more subtle in the case of dependent type since the later fields can have a type dependent on the former fields.  Consider a small example where we have consecutive fields \mintinline{Coq}{{.. a : nat := 1; a-def : Eq a 1 := eq_refl; ..}}, should we support override "a"? If we do support it, how can our compiler detect that "a-def" is not inheritable and signal an error when such thing happens?

@@

These two examples both show the incompatibility between exposing the concrete definition of a field for reasoning (this second example even involves the definition of fixpoint and pattern matching) and overridability. We need to be careful about this tension when designing the language.

\textlabel{Challenge (3)}{chg:consistency}~\textbf{Consistency vs. Self Parametrization}. The metatheoretical formulation in~\citet{zm2017} for family polymorphism is very natural as they consider an extension of the family as the concatenation of a new linkage. However, this formulation is not directly adaptable for dependent type theory because it can easily encode non-termination sabotaging consistency by the presence of "self". Informally, in the orignal formulation, we can override the "f" in family $\{"f" := id; "g" := "call self.f"\}$ with "call self.g", resulting the mutual calling of "f" and "g" thus a deadloop. This conflict between self parametrization and consistency requests changing the original formulation of family and inheritance for dependent type theory.

\textlabel{Challenge (4)}{chg:software-engineering}~\textbf{Engineering and Ecosystem}.  Not only for the metatheory, we also want our implementation of family to have good compatibility with the rest of Coq. For example, we shouldn't compromise \textbf{tactic programming} as it is part of the idiomatic Coq programming. Additionally, the interactive development style requires \textbf{incremental type checking}---our design has to be able to type check each Coq Vernacular command once inputted. What's more, we want our family ``term'' to incorporate Coq \textbf{without breaking its other functionality}. For example, the user may want to project the family member anywhere inside an arbitrary Gallina term, that may later be computed and normalized. However, families themselves can be second class citizens just like \citet{zm2017}.\YZ{I was thinking if the paper should start by talking about the functional encoding of inheritance (namely exposing a self parameter to allow late binding, and later taking the fixpoint of the extensible module when it needs to be accessed outside the family).  I think most readers will benefit from a recap of this encoding.\\
Coq is total but that doesn’t prevent Coq from having a Fixpoint keyword (plus you proved your language is consistent).
}
